### 2019-12-15 ic-journal/contraction.tex
Synchronized all changes from .tex to .htex; clean up all reviewed highlights (adding 3 new)

Milner's ``unique solution of equations'' theorem for $\wb$
(Theorem~\ref{t:Mil89})
brings a new proof technique for proving (weak) bisimilarities. However, it has
 limitations: the equations must be guarded and sequential. (\hl{Moreover,}
all sums where equation variables appear must be guarded sums.)
This limits the usefulness of the technique, since
the occurrences of other operators \hl{using equation} variables, such as parallel
composition and restriction,
in general cannot be \hl{eliminated}. % without changing the meaning of equations
The constraints \hl{in} Theorem~\ref{t:Mil89}, however, can be
weakened if we move from equations to a special kind of inequations called
  \emph{contractions}.

Intuitively, the bisimilarity contraction $\mcontrBIS$ is a preorder
\hlD{in which} $P \mcontrBIS Q$ holds if $P \wb Q$ and, in addition, 
\emph{$Q$ has the possibility of being at least as efficient as $P$} (as far as
$\tau$-actions are performed).
The process $Q$, however, may be nondeterministic and may have other ways
\hl{to do} the same work, \hl{ways} which could be \hl{slower} (i.e., involving
more $\tau$-actions than those performed by $P$).
% Thus, in contrast with expansion,  we cannot really say that `$Q$ is more efficient than
% $P$'.

@@

\begin{enumerate}
\item $P \arr\mu P'$ implies \hl{that} there is $Q'$ \hl{with} $Q \arcap \mu
  Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$ implies \hl{that} there is $P'$ \hl{with} $P \Arcap \mu
 P'$ and $P' \wb Q'$.
\end{enumerate}
Two processes $P$ and $Q$ are in the \emph{bisimilarity
contraction}, written as $P \mcontrBIS Q$,
if $P\ \R\ Q$ for some contraction $\R$.
Sometimes we write $\mexpaBIS$ for the inverse of $\mcontrBIS$.
\end{definition}
In clause (1) \hl{of the above definition}, $Q$ is required to match the challenge
transition \hl{of $P$} with at most one transition.
This makes sure that $Q$ is capable of mimicking % verb: mimic
$P$'s work at least as efficiently as $P$. 
In contrast, \hl{clause (2) entirely ignores efficiency on the challenges from $Q$:}
the final derivatives are required to be related by $\wb$, rather than by $\R$.

Bisimilarity contraction is coarser than bisimilarity expansion
$\expa$~\cite{arun1992efficiency,sangiorgi2015equations}, \hl{one of the
most useful auxiliary relations in up-to techniques}:
\begin{definition}[expansion]
\label{d:expa}
A process relation ${\R}$
  is an \emph{expansion} if, whenever $P\RR Q$,
 \begin{enumerate}
 \item   $P \arr\mu P'$ implies that there is $Q'$ with $Q \arcap \mu  Q'$
  and $P' \RR Q'$;
 \item $Q \arr\mu Q'$ implies that there is $P'$ with $P \Arr \mu P'$ and $P' \RR Q'$.
 \end{enumerate}
Two processes $P$ and $Q$ are in the \emph{bisimilarity
  expansion}, written as $P \expa Q$, if $P \RR Q$ for some expansion $\R$.
 \end{definition}
\hl{Bisimilarity expansion} is widely used in proof techniques for bisimilarity.
\hl{It} intuitively refines bisimilarity by 
formalising the idea of ``efficiency'' between processes.
Clause (1) is the same in the both preorders, while in clause (2) expansion \hl{requires}
$P \Arr \mu P'$, rather than $P \Arcap \mu P'$.
\hl{Moreover,} in clause (2) of Def.~\ref{d:BisCon} the final derivatives
are simply required to be bisimilar ($P' \wb Q'$).
Intuitively, $P \expa Q$ holds if $P\wb Q$ and, in addition, \emph{$Q$
  is always at least as efficient as $P$}.

@@

\hlD{Bisimilarity expansion and bisimilarity contraction are both
preorders.}
Similarily with (weak) bisimilarity, both the expansion and the
contraction preorders are preserved by all CCS operators except the
summation. The proofs are similar \hlD{to those for bisimilarity},
see, e.g.~\cite{sangiorgi2017equations} \hl{for details.}

>>>>

Milner's ``unique solution of equations'' theorem for $\wb$
(Theorem~\ref{t:Mil89})
brings a new proof technique for proving (weak) bisimilarities. However, it has
 limitations: the equations must be guarded and sequential. (Moreover,
all sums where equation variables appear must be guarded sums.)
This limits the usefulness of the technique, since
the occurrences of other operators using equation variables, such as parallel
composition and restriction,
in general cannot be eliminated. % without changing the meaning of equations
The constraints in Theorem~\ref{t:Mil89}, however, can be
weakened if we move from equations to a special kind of inequations called
  \emph{contractions}.

Intuitively, the bisimilarity contraction $\mcontrBIS$ is a preorder
in which $P \mcontrBIS Q$ holds if $P \wb Q$ and, in addition, 
\emph{$Q$ has the possibility of being at least as efficient as $P$} (as far as
$\tau$-actions are performed).
The process $Q$, however, may be nondeterministic and may have other ways
to do the same work, ways which could be slower (i.e., involving
more $\tau$-actions than those performed by $P$).
% Thus, in contrast with expansion,  we cannot really say that `$Q$ is more efficient than
% $P$'.

@@

\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ with $Q \arcap \mu
  Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$ implies that there is $P'$ with $P \Arcap \mu
 P'$ and $P' \wb Q'$.
\end{enumerate}
Two processes $P$ and $Q$ are in the \emph{bisimilarity
contraction}, written as $P \mcontrBIS Q$,
if $P\ \R\ Q$ for some contraction $\R$.
Sometimes we write $\mexpaBIS$ for the inverse of $\mcontrBIS$.
\end{definition}
In clause (1) of the above definition, $Q$ is required to match the challenge
transition of $P$ with at most one transition.
This makes sure that $Q$ is capable of mimicking % verb: mimic
$P$'s work at least as efficiently as $P$. 
In contrast, clause (2) entirely ignores efficiency on the challenges from $Q$:
the final derivatives are required to be related by $\wb$, rather than by $\R$.

Bisimilarity contraction is coarser than bisimilarity expansion
$\expa$~\cite{arun1992efficiency,sangiorgi2015equations}, one of the
most useful auxiliary relations in up-to techniques:
\begin{definition}[expansion]
\label{d:expa}
A process relation ${\R}$
  is an \emph{expansion} if, whenever $P\RR Q$,
 \begin{enumerate}
 \item   $P \arr\mu P'$ implies that there is $Q'$ with $Q \arcap \mu  Q'$
  and $P' \RR Q'$;
 \item $Q \arr\mu Q'$ implies that there is $P'$ with $P \Arr \mu P'$ and $P' \RR Q'$.
 \end{enumerate}
Two processes $P$ and $Q$ are in the \emph{bisimilarity
  expansion}, written as $P \expa Q$, if $P \RR Q$ for some expansion $\R$.
 \end{definition}
Bisimilarity expansion is widely used in proof techniques for bisimilarity.
It intuitively refines bisimilarity by 
formalising the idea of ``efficiency'' between processes.
Clause (1) is the same in the both preorders, while in clause (2) expansion requires
$P \Arr \mu P'$, rather than $P \Arcap \mu P'$.
Moreover, in clause (2) of Def.~\ref{d:BisCon} the final derivatives
are simply required to be bisimilar ($P' \wb Q'$).
Intuitively, $P \expa Q$ holds if $P\wb Q$ and, in addition, \emph{$Q$
  is always at least as efficient as $P$}.

@@

Bisimilarity expansion and bisimilarity contraction are both preorders.
Similarily with (weak) bisimilarity, both the expansion and the
contraction preorders are preserved by all CCS operators except the
summation. The proofs are similar to those for bisimilarity,
see, e.g.~\cite{sangiorgi2017equations} for details.

### 2019-12-15 ic-journal/equa.tex
Synchronized all changes from .tex to .htex; clean up all reviewed highlights (adding 3 new)

In this paper (and the formalisation work), we use the agent
variables also as \emph{equation variables}. This eliminates the need of
\hl{an additional} type for \hl{CCS} equations, and we can reuse the existing
variable substitution operation (cf.~the SOS rule for the Recursion in
Fig.~\ref{f:LTSCCS}) for \hl{substitutions} of equation variables.
For example, the result of substituting  variable $X$ with $\nil$ in $a.X +
\recu X (b.X)$,  written \hl{as} $(a.X + \recu X (b.X)) \sub {\nil} X$, is
$a.\nil + \recu X (b.X)$ with $\recu X (b.X)$
untouched. \Multivariate substitutions are written in the same syntax,
e.g. $E \sub {\til P} {\til X}$. Whenever $\til X$ is clear from the
context, we may also write $E[\til P]$ instead of $E \sub {\til P} {\til
  X}$ (and $E[P]$ for $E \sub P X$ if there is a single equation
variable $X$).

@@

\begin{definition}[solutions and the unique solution]
  \label{def:solution}
Suppose $\{ X_i = E_i\}_{i\in I}$ is a system of equations: 
\begin{itemize}
\item
 $\til P$ is a \emph{solution of the system of equations (for $\wb$)} 
if for each $i$ it holds that $P_i \wb E_i [\til P]$;
\item The system has \emph{a unique solution for $\wb$}  if whenever 
 $\til P$ and $\til Q$ are both solutions then $\til P \wb \til Q$. 
\end{itemize} 
 \end{definition}
Similarily, the \emph{(unique) solution of a system of equations for $\sim$}
(or for $\rapprox$) can be obtained by replacing all occurrences of $\wb$
in \hl{the} above definition with $\sim$ and $\rapprox$, respectively.

@@

The version of Milner's unique-solution theorem for $\wb$ further requires
that all sums are guarded:
% \footnote{But if the CCS syntax were defined with only guarded
%   sums, i.e., $\sum_{i\in I} \mu_i.P_i$ as
%   in~\cite{sangiorgi2015equations}, this addition
%   requirement disappears automatically, and we can even say $\wb$ is indeed a
%   congruence.}:
\begin{theorem}[unique solution of equations for $\wb$]
\label{t:Mil89}
Let $E_i$   \hl{(with only guarded sums)}  be guarded and sequential,
and  with free
variables in $\til X$. Let ${\til P} \wb {\til E}\{\til P /\til X\}$,
  ${\til Q} \wb {\til E}\{\til Q /\til X\}$, then ${\til P} \wb {\til Q}$.
\end{theorem}

The proof of the last two theorems above exploits an invariance
property on immediate derivatives
of guarded and sequential expressions, and then extracts a bisimulation
(up to bisimilarity) out of the solutions of the system.
To see the need of the sequentiality condition, consider
 the equation (from~\cite{Mil89}) $X \wb \res a (a. X | \outC a)$
where $X$ is guarded but not sequential. Any process that does not perform
$a$\hl{--action} is a solution, e.g. $\nil$ and $b.\nil$.

For more details \hl{on the formalisation of the} above three theorems, see Section~\ref{ss:part2}
for the \univariate case and Section~\ref{sec:multivariate} for the \multivariate case.

>>>>

In this paper (and the formalisation work), we use the agent
variables also as \emph{equation variables}. This eliminates the need of
an additional type for CCS equations, and we can reuse the existing
variable substitution operation (cf.~the SOS rule for the Recursion in
Fig.~\ref{f:LTSCCS}) for substitutions of equation variables.
For example, the result of substituting  variable $X$ with $\nil$ in $a.X +
\recu X (b.X)$,  written as $(a.X + \recu X (b.X)) \sub {\nil} X$, is
$a.\nil + \recu X (b.X)$ with $\recu X (b.X)$
untouched. \Multivariate substitutions are written in the same syntax,
e.g. $E \sub {\til P} {\til X}$. Whenever $\til X$ is clear from the
context, we may also write $E[\til P]$ instead of $E \sub {\til P} {\til
  X}$ (and $E[P]$ for $E \sub P X$ if there is a single equation
variable $X$).

@@

\begin{definition}[solutions and the unique solution]
  \label{def:solution}
Suppose $\{ X_i = E_i\}_{i\in I}$ is a system of equations: 
\begin{itemize}
\item
 $\til P$ is a \emph{solution of the system of equations (for $\wb$)} 
if for each $i$ it holds that $P_i \wb E_i [\til P]$;
\item The system has \emph{a unique solution for $\wb$}  if whenever 
 $\til P$ and $\til Q$ are both solutions then $\til P \wb \til Q$. 
\end{itemize} 
 \end{definition}
Similarily, the \emph{(unique) solution of a system of equations for $\sim$}
(or for $\rapprox$) can be obtained by replacing all occurrences of $\wb$
in the above definition with $\sim$ and $\rapprox$, respectively.

@@

The version of Milner's unique-solution theorem for $\wb$ further requires
that all sums are guarded:
% \footnote{But if the CCS syntax were defined with only guarded
%   sums, i.e., $\sum_{i\in I} \mu_i.P_i$ as
%   in~\cite{sangiorgi2015equations}, this addition
%   requirement disappears automatically, and we can even say $\wb$ is indeed a
%   congruence.}:
\begin{theorem}[unique solution of equations for $\wb$]
\label{t:Mil89}
Let $E_i$ (with only guarded sums) be guarded and sequential,
and  with free
variables in $\til X$. Let ${\til P} \wb {\til E}\{\til P /\til X\}$,
  ${\til Q} \wb {\til E}\{\til Q /\til X\}$, then ${\til P} \wb {\til Q}$.
\end{theorem}

The proof of the last two theorems above exploits an invariance
property on immediate derivatives
of guarded and sequential expressions, and then extracts a bisimulation
(up to bisimilarity) out of the solutions of the system.
To see the need of the sequentiality condition, consider
 the equation (from~\cite{Mil89}) $X \wb \res a (a. X | \outC a)$
where $X$ is guarded but not sequential. Any process that does not perform
$a$--action is a solution, e.g. $\nil$ and $b.\nil$.

For more details on the formalisation of the above three theorems, see Section~\ref{ss:part2}
for the \univariate case and Section~\ref{sec:multivariate} for the \multivariate case.

### 2019-12-15 ic-journal/expansion.tex
Synchronized all changes from .tex to .htex; clean up all reviewed highlights (adding 3 new)

\emph{Strong} bisimulation and strong bisimilarity ($\sim$)
can be obtained by replacing the weak transition $Q\Arcap\mu Q'$ in 
clause (1) 
 with the transition $Q \arr \mu Q'$ (and similarly for the other clause).
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums), i.e.~$P_1 \wb Q_1$ and $P_2 \wb Q_2$  \hl{do} not imply
 $P_1 + P_2 \wb Q_1 + Q_2$.
For this \hl{reason}, Milner introduced the concept of
\emph{observational congruence}, also called \emph{rooted
  bisimilarity}~\cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \emph{rooted bisimilar}, written as $P
\rapprox Q$, if % for all $\mu$:
%  for all $\mu\in \mathscr{L}\cup\{\tau\}$
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}
% This new sentence maintains a link with Definition 2.1
Relation $\rapprox$ is indeed preserved by the sum operator.
% \hl{It can be proved that $P_1 \rapprox Q_1 \wedge P_2 \rapprox Q_2 \Longrightarrow
% P_1 + P_2 \rapprox Q_1 + Q_2$.}
The above definition also brings up a proof technique for proving rooted
bisimilarity from a given bisimulation. The next
lemma plays an
important role in proving some key results in this paper:
\begin{lemma}[rooted bisimilarity by bisimulation]
\label{l:obsCongrByWeakBisim}
Given a (weak) bisimulation $\RR$, suppose two processes $P$ and $Q$
satisfy the following properties:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \RR Q'$.
\end{enumerate}
Then $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

One basic property of rooted bisimiarity is congruence,
indeed it is  the coarsest congruence contained in bisimilarity
(see Section~\ref{s:coarsest} and \hl{\mbox{Section~\ref{ss:context}}} for more details):
\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\wb$.
\end{theorem}

>>>>

\emph{Strong} bisimulation and strong bisimilarity ($\sim$)
can be obtained by replacing the weak transition $Q\Arcap\mu Q'$ in 
clause (1) 
 with the transition $Q \arr \mu Q'$ (and similarly for the other clause).
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums), i.e.~$P_1 \wb Q_1$ and $P_2 \wb Q_2$ do not imply
 $P_1 + P_2 \wb Q_1 + Q_2$.
For this reason, Milner introduced the concept of
\emph{observational congruence}, also called \emph{rooted
  bisimilarity}~\cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \emph{rooted bisimilar}, written as $P
\rapprox Q$, if % for all $\mu$:
%  for all $\mu\in \mathscr{L}\cup\{\tau\}$
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}
% This new sentence maintains a link with Definition 2.1
Relation $\rapprox$ is indeed preserved by the sum operator.
The above definition also brings up a proof technique for proving rooted
bisimilarity from a given bisimulation. The next
lemma plays an
important role in proving some key results in this paper:
\begin{lemma}[rooted bisimilarity by bisimulation]
\label{l:obsCongrByWeakBisim}
Given a (weak) bisimulation $\RR$, suppose two processes $P$ and $Q$
satisfy the following properties:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \RR Q'$.
\end{enumerate}
Then $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

One basic property of rooted bisimiarity is congruence,
indeed it is  the coarsest congruence contained in bisimilarity
(see Section~\ref{s:coarsest} and Section~\ref{ss:context} for more details):
\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\wb$.
\end{theorem}

### 2019-12-15 ic-journal/formal.tex
Synchronized all changes from .tex to .htex; clean up all reviewed highlights (adding 3 new)

We highlight here a \hl{comprehensive} formalisation of CCS in the HOL theorem prover
(HOL4)~\cite{Melham:1993vl,slind2008brief}, with a focus towards the
theory (and formal proofs) of the unique solution of
equations/contractions theorems mentioned in Section~\ref{s:eq} and
\ref{s:mcontr}.
All proof scripts are available as part of HOL's official
examples\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL/tree/master/examples/CCS}}.
The work so far consists of about 24,000 lines (1MB) of code in total,
in which about 5,000 lines were derived from the early work of Monica
Nesi~\cite{Nesi:1992ve} on HOL88, with major modifications.

Higher Order Logic (HOL)~\cite{hollogic} traces
its roots back to the
\emph{Logic of Computable Functions
  (LCF)}~\cite{gordon1979edinburgh,milner1972logic}
by Robin Milner and others
since 1972. It is a variant of
Church’s Simple Theory of Types (STT)~\cite{church1940formulation},
plus a higher order version of Hilbert's choice operator $\varepsilon$,
Axiom of Infinity, and Rank-1 (prenex) polymorphism.
HOL4 has implemented the original HOL, 
while some other theorem provers in the HOL family (e.g.~Isabelle/HOL) have
certain extensions.
%  (they made the formal language more powerful,
% but they also bring the possibilities that the entire logic becomes
% inconsistent). 
Indeed, HOL has considerably simpler logical
foundations than most other theorem provers.
As a consequence, theories and proofs verified in HOL are easier to understand
 \hl{for} people who are not familar with more advanced dependent type
 theories, \hl{e.g. the Calculus of Constructions implemented in Coq.}

@@

\hl{HOL terms} represent elements of the
sets denoted by \hl{their} types. There are four kinds of \hl{HOL terms,
which} can be described (in simplified forms) by the BNF grammar in
Fig.~\ref{fig:hol-terms}. (See~\cite{hollogic} for a complete
description of HOL, \hl{including the primitive derivative rules to be
mentioned below.})

>>>>

We highlight here a comprehensive formalisation of CCS in the HOL theorem prover
(HOL4)~\cite{Melham:1993vl,slind2008brief}, with a focus towards the
theory (and formal proofs) of the unique solution of
equations/contractions theorems mentioned in Section~\ref{s:eq} and
\ref{s:mcontr}.
All proof scripts are available as part of HOL's official
examples\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL/tree/master/examples/CCS}}.
The work so far consists of about 24,000 lines (1MB) of code in total,
in which about 5,000 lines were derived from the early work of Monica
Nesi~\cite{Nesi:1992ve} on HOL88, with major modifications.

Higher Order Logic (HOL)~\cite{hollogic} traces
its roots back to the
\emph{Logic of Computable Functions
  (LCF)}~\cite{gordon1979edinburgh,milner1972logic}
by Robin Milner and others
since 1972. It is a variant of
Church’s Simple Theory of Types (STT)~\cite{church1940formulation},
plus a higher order version of Hilbert's choice operator $\varepsilon$,
Axiom of Infinity, and Rank-1 (prenex) polymorphism.
HOL4 has implemented the original HOL, 
while some other theorem provers in the HOL family (e.g.~Isabelle/HOL) have
certain extensions.
%  (they made the formal language more powerful,
% but they also bring the possibilities that the entire logic becomes
% inconsistent). 
Indeed, HOL has considerably simpler logical
foundations than most other theorem provers.
As a consequence, theories and proofs verified in HOL are easier to understand
 for people who are not familar with more advanced dependent type
 theories, e.g. the Calculus of Constructions implemented in Coq.

@@

HOL terms represent elements of the
sets denoted by their types. There are four kinds of HOL terms,
which can be described (in simplified forms) by the BNF grammar in
Fig.~\ref{fig:hol-terms}. (See~\cite{hollogic} for a complete
description of HOL, including the primitive derivative rules to be
mentioned below.)

### 2019-12-15 ic-journal/intro.tex
Synchronized all changes from .tex to .htex; clean up all reviewed highlights (adding 3 new)

Milner's unique-solution theorem for weak bisimilarity, however,
has severe syntactic limitations:
the equations must be both \emph{guarded} and \emph{sequential}. That is,
the variables of the equations can only occur underneath visible prefixes and \hl{summation}.
One way to overcome \hl{these} limitations is to replace the equations
with special inequations called
\emph{contractions}~\cite{sangiorgi2015equations,sangiorgi2017equations}. \hl{The
contraction relation} is a
preorder that, roughly, places some efficiency constraints on processes.
The unique solution of contractions is defined as with equations:
any two solutions must be componentwise bisimilar.
The difference from equations is in the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem are
eliminated. One constraint that still remains
in~\cite{sangiorgi2017equations}
(where the issue is bypassed \hl{by} using a more
restrictive CCS syntax)
is the occurrences of \hlD{arbitrary} sums (e.g.~$P + Q$) due to
the non-substitutivity of the contraction preorder in this case.

This paper presents a comprehensive formalisation of Milner's Calculus of Communicating
Systems (CCS) in the HOL theorem prover (HOL4),
with a focus towards the theory of unique solutions of equations and contractions.
Many results in Milner's CCS book~\cite{Mil89} are covered, since
the unique-solution theorems rely on a large number of fundamental results.
Indeed the formalisation encompasses all basic properties of strong,
\hl{weak and rooted}
bisimilarities (e.g.~the fixed-point and substitutivity properties), \hl{and} their algebraic laws.
Further extensions include several versions of ``bisimulation up to''
techniques, \hl{and properties of the expansion and contraction preorders}.
Concerning rooted bisimilarity, the formalisation
includes Hennessy's Lemma, Deng's Lemma,
 and two theorems saying that rooted bisimilarity is the coarsest (largest)
 congruence contained in weak bisimilarity ($\wb$): one is classical
 with the hypothesis that no process uses up all labels;
the other one is 
without such hypothesis, essentially \hl{formalising} van Glabbeek's proof~\cite{van2005characterisation}.
\hl{With} this respect, the work is also an extensive experiment using the HOL theorem prover with its
recent developments, including its coinduction package.

This formalisation has offered us the possibility of
further refining the theory of unique solutions of contractions.
In particular,  existing results~\cite{sangiorgi2017equations} have
limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.
In this paper, the contraction-based proof technique is further
refined by moving from the contraction preorder to
\emph{rooted contraction}, \hl{which is shown to be} the coarsest
precongruence contained in the contraction preorder.
The resulting unique-solution theorem is now valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
constraints on \hl{summation}.

Another benefit of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures.
Such structural simililarities can be found, for instance, in the
following cases: the proofs that rooted bisimilarity and rooted
contraction are, respectively, the coarsest congruence contained in
weak bisimilarity and the coarsest precongruence contained in the
contraction preorder; the proofs about unique solution(s) of equations
for weak bisimilarity that use the contraction preorder as an
auxiliary relation, and other \hl{unique-solution theorems} (e.g. the one
for rooted bisimilarity 
 in which the auxiliary relation is rooted contraction); the
proofs about various forms of enhancements of the bisimulation proof
method (the ``up-to'' techniques).
%
In these cases, when moving between proofs there are only a few places \hl{in}
the HOL proof scripts \hlD{that have} to be modified.
Then the successful termination of a proof gives us the
 guarantee that the proof is \hl{correct}, eliminating the risks 
of overlooking or missing details as in \emph{paper-and-pencil} proofs.

\hl{Concerning the formalisation of unique-solution theorems,}
we first consider the case of a single equation (or contraction), the
\emph{\univariate case}.
\hl{Then we turn to} the \emph{\multivariate case} where more equations
  (or contractions) with multiple equation variables \hl{are involved}.
  The \univariate versions \hl{of the} unique-solution theorems
  \hl{can directly use}
$\lambda$-functions to \hl{represent} CCS equations,
while the \multivariate versions \hl{require}
more careful and delicate \hl{treatments} of CCS
expressions, with multiple variables and substitutions acting on them.
% should be in the same paragraph, still talking about multivariable equations.
In contrast to literature such as~\cite{Gorrieri:2015jt},
we have followed Milner's original approach~\cite{milner1990operational} and adopted the same
type for both CCS equations and processes: those undefined constants
in CCS terms are treated as  (free) equation variables, and a CCS
process is a CCS \hl{expression} without equation variables.  % don't use "term"
This allows us a smoother move from the \univariate case to the \multivariate
one. (See Section~\ref{sec:multivariate} for more details.)

@@

In Section~\ref{ss:ccs} \hl{we recall the core theory} of CCS,
including its syntax, operational semantics, bisimilarity and rooted
bisimilarity.
\hl{Then,} Section~\ref{s:eq} discusses equations and contractions, and in particular,
Section~\ref{ss:new} presents rooted contraction and the related
unique-solution result for rooted bisimilarity.
In Section~\ref{s:for} we highlight the CCS formalisation in HOL4, with
the unique-solution theorems in the \univariate case.
Section~\ref{sec:multivariate} describes the extension to the \multivariate case.
Finally, in Section~\ref{s:rel} and \ref{s:concl} \hl{we discuss} the related work,
conclusions, and a few directions for future work.

>>>>

Milner's unique-solution theorem for weak bisimilarity, however,
has severe syntactic limitations:
the equations must be both \emph{guarded} and \emph{sequential}. That is,
the variables of the equations can only occur underneath visible prefixes and summation.
One way to overcome these limitations is to replace the equations
with special inequations called
\emph{contractions}~\cite{sangiorgi2015equations,sangiorgi2017equations}. The
contraction relation is a
preorder that, roughly, places some efficiency constraints on processes.
The unique solution of contractions is defined as with equations:
any two solutions must be componentwise bisimilar.
The difference from equations is in the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem are
eliminated. One constraint that still remains
in~\cite{sangiorgi2017equations}
(where the issue is bypassed by using a more
restrictive CCS syntax)
is the occurrences of arbitrary sums (e.g.~$P + Q$) due to
the non-substitutivity of the contraction preorder in this case.

This paper presents a comprehensive formalisation of Milner's Calculus of Communicating
Systems (CCS) in the HOL theorem prover (HOL4),
with a focus towards the theory of unique solutions of equations and contractions.
Many results in Milner's CCS book~\cite{Mil89} are covered, since
the unique-solution theorems rely on a large number of fundamental results.
Indeed the formalisation encompasses all basic properties of strong,
weak and rooted
bisimilarities (e.g.~the fixed-point and substitutivity properties), and their algebraic laws.
Further extensions include several versions of ``bisimulation up to''
techniques, and properties of the expansion and contraction preorders.
Concerning rooted bisimilarity, the formalisation
includes Hennessy's Lemma, Deng's Lemma,
 and two theorems saying that rooted bisimilarity is the coarsest (largest)
 congruence contained in weak bisimilarity ($\wb$): one is classical
 with the hypothesis that no process uses up all labels;
the other one is 
without such hypothesis, essentially formalising van Glabbeek's proof~\cite{van2005characterisation}.
With this respect, the work is also an extensive experiment using the HOL theorem prover with its
recent developments, including its coinduction package.

This formalisation has offered us the possibility of
further refining the theory of unique solutions of contractions.
In particular,  existing results~\cite{sangiorgi2017equations} have
limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.
In this paper, the contraction-based proof technique is further
refined by moving from the contraction preorder to
\emph{rooted contraction}, which is shown to be the coarsest
precongruence contained in the contraction preorder.
The resulting unique-solution theorem is now valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
constraints on summation.

Another benefit of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures.
Such structural simililarities can be found, for instance, in the
following cases: the proofs that rooted bisimilarity and rooted
contraction are, respectively, the coarsest congruence contained in
weak bisimilarity and the coarsest precongruence contained in the
contraction preorder; the proofs about unique solution(s) of equations
for weak bisimilarity that use the contraction preorder as an
auxiliary relation, and other unique-solution theorems (e.g. the one
for rooted bisimilarity 
 in which the auxiliary relation is rooted contraction); the
proofs about various forms of enhancements of the bisimulation proof
method (the ``up-to'' techniques).
%
In these cases, when moving between proofs there are only a few places in
the HOL proof scripts that have to be modified.
Then the successful termination of a proof gives us the
 guarantee that the proof is correct, eliminating the risks 
of overlooking or missing details as in \emph{paper-and-pencil} proofs.

Concerning the formalisation of unique-solution theorems,
we first consider the case of a single equation (or contraction), the
\emph{\univariate case}.
Then we turn to the \emph{\multivariate case} where more equations
  (or contractions) with multiple equation variables are involved.
  The \univariate versions of the unique-solution theorems
  can directly use
$\lambda$-functions to represent CCS equations,
while the \multivariate versions require
more careful and delicate treatments of CCS
expressions, with multiple variables and substitutions acting on them.
% should be in the same paragraph, still talking about multivariable equations.
In contrast to literature such as~\cite{Gorrieri:2015jt},
we have followed Milner's original approach~\cite{milner1990operational} and adopted the same
type for both CCS equations and processes: those undefined constants
in CCS terms are treated as  (free) equation variables, and a CCS
process is a CCS expression without equation variables.  % don't use "term"
This allows us a smoother move from the \univariate case to the \multivariate
one. (See Section~\ref{sec:multivariate} for more details.)

@@

In Section~\ref{ss:ccs} we recall the core theory of CCS,
including its syntax, operational semantics, bisimilarity and rooted
bisimilarity.
Then, Section~\ref{s:eq} discusses equations and contractions, and in particular,
Section~\ref{ss:new} presents rooted contraction and the related
unique-solution result for rooted bisimilarity.
In Section~\ref{s:for} we highlight the CCS formalisation in HOL4, with
the unique-solution theorems in the \univariate case.
Section~\ref{sec:multivariate} describes the extension to the \multivariate case.
Finally, in Section~\ref{s:rel} and \ref{s:concl} we discuss the related work,
conclusions, and a few directions for future work.
